Ferromagnetism and spin excitations in topological Hubbard models with a flat-band

### 2018-06-26 paper/paper.tex
Version 1.2.

\begin{document}
\title{Flatband ferromagnetism and dispersion relations of spin-1 excitations in topological Hubbard models}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\email[]{jxli@nju.edu.cn}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in topological Hubbard models. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction strength upon which the ferromagnetic phase is stable, and elaborate the ferromagnetic spin-1 excitation spectra of the Chern Hubbard model and the $Z_2$ Hubbard model. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase.
\end{abstract}
\maketitle

@@

\par The interplay between flatband ferromagnetism and nontrivial band topology enriches the related physics. In fact, ferromagnetism is essential in the generation of stable fractional Chern insulators in the proposals where the spin degrees of freedom of electrons are considered \cite{TMW_PRL2011,NSRCM_PRB2011,LBFL_PRL2012}. Furthermore, ferromagnetism can also lead to possible high-temperature quantum anomalous Hall effect (QAHE) when the nearly-flat topological band is half-filled \cite{NSRCM_PRL2012}. In this paper, we focus on this ferromagnetism induced QAHE in nearly-flat topological bands. As a paradigm, we consider a square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). When the model is quarter filled (or correspondingly, the lower nearly-flat band is half filled), the ground state is spin fully polarized due to the ferromagnetism and exhibits QAHE because of the nonzero Chern number of a single-spin band. Then the low energy physics is dominated by the one-spin-flip excitations. These spin-1 excitations has been studied by a generalized bosonization scheme where the interacting fermionic model is mapped to a free bosonic model describing spin-wave excitations at the harmonic approximation \cite{DG_PRB2015}. The ferromagnetism was shown to be stable against such spin wave excitations, which are gapless in the Chern Hubbard model and gapped in the $Z_2$ Hubbard model. However in this bosonization scheme, the free part of the electron model plays no role in the spin wave excitations other than contributing a global constant, which means it should fail due to the competition between the kinetic energy and potential energy of electrons when the nonflatness of the electron bands is not negligible \cite{T_PRL1994}. What's worse, in a strictly local periodic tight-binding model, an energy band with a nonzero Chern number cannot be exactly flat \cite{CMST_JPMT2014}. Therefore it remains an open question on whether the ground state is stable against the spin-1 excitations and how the the nonflatness of the electron bands manifests itself in the spin-1 excitation spectra in such models.

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical Hubbard interaction strength is found for both the Chern Hubbard model and the $Z_2$ Hubbard model, upon which the ferromagnetic phase is stable. Furthermore, the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase.

@@

\par We want to make a comparison of our flatband-limit results with those obtained by the generalized bosonization formalism proposed in Ref. \cite{DG_PRB2015}, where both the nonflatness of the electron bands and the self-interactions between the magnons are ignored. For the Chern Hubbard model, our results are qualitatively consistent with theirs [See Fig. \ref{fmcispectrum}(a) in our paper and Fig. 4 in Ref. \cite{DG_PRB2015}]. However, for the $Z_2$ Hubbard model, our results are different from theirs in that the band bottom of the spin waves in our results lies at the $M_{1,2,3,4}$ points while theirs lies at the $\Gamma$ point [See Figs. \ref{tispectrum}(a)-(b) in our paper and Fig. 5 in Ref. \cite{DG_PRB2015}]. We attribute this discrepancy to the self-interactions of the magnons because our numerical method treat the system after the projection exactly. Although correlated nearly-flat topological bands are hard to realize in real materials, they are possible to be designed in cold-atom systems. In fact great progress has been made very recently \cite{LCJPS_N2009,AALBPB_PRL2013,MSKBK_PRL2013,JMDLUGE_N2014}. We thus expect our results to be tested in future cold-atom experiments.

>>>>

\begin{document}
\title{Ferromagnetism and spin excitations in topological Hubbard models with a flat-band}
\author{Xiao-Fei Su}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{School of Physics and Electronic Information, Huaibei Normal University, Huaibei 235000, China}
\author{Zhao-Long Gu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Zhao-Yang Dong}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\author{Shun-Li Yu}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\author{Jian-Xin Li}
\email[]{jxli@nju.edu.cn}
\affiliation{National Laboratory of Solid State Microstructures and Department of Physics, Nanjing University, Nanjing 210093, China}
\affiliation{Collaborative Innovation Center of Advanced Microstructures, Nanjing University, Nanjing 210093, China}
\date{\today}

\begin{abstract}
\par We study the spin-1 excitation spectra of the flatband ferromagnetic phases in topological Hubbard models. As a paradigm, we consider a quarter filled square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). By using the numerical exact diagonalization method with a projection onto the lower nearly-flat electron band, we determine the critical Hubbard interaction upon which the ferromagnetic phase is stable, and obtain the ferromagnetic spin-1 excitation spectra of the Chern Hubbard model and the $Z_2$ Hubbard model. Both spectra consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, in both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which can lead to roton-like spin wave excitations. We elaborate that the destabilization of the ferromagnetic phase happens when the energy of the induced roton-like modes becomes zero with the increase of the nonflatness.
\end{abstract}
\maketitle

@@

\par The interplay between flatband ferromagnetism and nontrivial band topology enriches the related physics. In fact, ferromagnetism is essential in the generation of stable fractional Chern insulators in the proposals where the spin degrees of freedom of electrons are considered \cite{TMW_PRL2011,NSRCM_PRB2011,LBFL_PRL2012}. Furthermore, ferromagnetism can also lead to possible high-temperature quantum anomalous Hall effect (QAHE) when the nearly-flat topological band is half-filled \cite{NSRCM_PRL2012}. In this paper, we focus on this ferromagnetism induced QAHE in nearly-flat topological bands. As a paradigm, we consider a square lattice Hubbard model whose free part is the $\pi$ flux model with topologically nontrivial and nearly-flat electron bands. The model Hamiltonian either explicitly breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry (Chern Hubbard model), or preserves the time-reversal symmetry but explicitly breaks the spin $SU(2)$ rotation symmetry ($Z_2$ Hubbard model). When the model is quarter filled (or correspondingly, the lower nearly-flat band is half filled), the ground state is spin fully polarized due to the ferromagnetism and exhibits QAHE because of the nonzero Chern number of a single-spin band. Then the low energy physics is dominated by the one-spin-flip excitations. These spin-1 excitations have been studied by a generalized bosonization scheme where the interacting fermionic model is mapped to a free bosonic model describing spin-wave excitations at the harmonic approximation \cite{DG_PRB2015}. The ferromagnetism was shown to be stable against such spin wave excitations, which are gapless in the Chern Hubbard model and gapped in the $Z_2$ Hubbard model. However in this bosonization scheme, the free part of the electron model plays no role in the spin wave excitations other than contributing a global constant, which means it should fail due to the competition between the kinetic energy and potential energy of electrons when the nonflatness of the electron bands is not negligible \cite{T_PRL1994}. In fact, in a strictly local periodic tight-binding model, an energy band with a nonzero Chern number cannot be exactly flat \cite{CMST_JPMT2014}. Therefore it remains an open question on whether the ground state is stable against the spin-1 excitations and how the nonflatness of the electron bands manifests itself in the spin-1 excitation spectra in such models.

\par To elucidate these questions, we adopt the numerical exact diagonalization method with a projection onto the lower nearly-flat band to take close investigations on the spin-1 excitations of the models. A critical magnitude of the Hubbard interaction is found for both the Chern Hubbard model and the $Z_2$ Hubbard model, upon which the ferromagnetic phase is stable. Furthermore, the spin-1 excitation spectra are shown to consist of collective modes (spin waves) and individual modes (Stoner continuum). For the Chern Hubbard model, the spin wave is gapless while for the $Z_2$ Hubbard model, the spin wave is gapped. Remarkably, for both cases, the nonflatness of the free electron bands introduces dips of the lower boundary of the Stoner continuum, and significantly renormalizes the energies of the collective modes around these dips downward, which leads to roton-like spin wave excitations. With the increase of this nonflatness, the energy of the induced roton-like modes goes down and finally touches zero, which results in the destabilization of the ferromagnetic phase. We also make a comparison of our results in the flatband limit with those obtained by the bosonization scheme \cite{DG_PRB2015}. Qualitative agreements are observed for the Chern Hubbard model, yet notable discrepancy appears for the $Z_2$ Hubbard model. We attribute the discrepancy to the self-interactions of the magnons, which is ignored in the bosonization scheme but treated exactly in our method after the projection.

@@

\par We want to make a comparison of our flatband-limit results with those obtained by the generalized bosonization formalism proposed in Ref. \cite{DG_PRB2015}, where both the nonflatness of the electron bands and the self-interactions between the magnons are ignored. For the Chern Hubbard model, our results are qualitatively consistent with theirs [See Fig. \ref{fmcispectrum}(a) in our paper and Fig. 4 in Ref. \cite{DG_PRB2015}]. However, for the $Z_2$ Hubbard model, our results are different from theirs at two points at least [See Figs. \ref{tispectrum}(a)-(b) in our paper and Fig. 5 in Ref. \cite{DG_PRB2015}]. On the one hand, the band bottom of the spin waves in our results lies at the $M_{1,2,3,4}$ points while theirs lies at the $\Gamma$ point . On the other hand, the acoustic branch and optical branch are degenerate along the $\Gamma$-$X$ path in our results while they are degenerate along the $X$-$M$ path in theirs. We attribute this discrepancy to the self-interactions of the magnons because our numerical method treat the system after the projection exactly. Although correlated nearly-flat topological bands are hard to realize in real materials, they are possible to be designed in cold-atom systems. In fact great progress has been made very recently \cite{LCJPS_N2009,AALBPB_PRL2013,MSKBK_PRL2013,JMDLUGE_N2014}. We thus expect our results to be tested in future cold-atom experiments.
